---
title: "Exercises"
output: html_document
---
## Exercises: Reading data, navigating through the data frame, applying descriptive statistics functions

Exercise 1: First, read the UNdata csv file into a data frame called UNdataExercise. Use the first column for row names; assume that the csv file has headers for the columns in the first row.
```{r}
UNdataExercise <- read.csv(file = 'UNdata.csv', row.names = 1)
```

Exercise 2: Using a single command, display the job satisfaction (SAT_JOB) and per capita GDP (GDP_PC) for the first three rows.
```{r}
UNdataExercise[1:3, c('SAT_JOB','GDP_PC')]
```

Exercise 3: Using a single command, determine the values for the 4th and 14th columns of the 3rd and 50th rows.
```{r}
UNdataExercise[c(3,50),c(4,14)]
```

Exercise 4: Use a single command to return the HDI of the country with the greatest education (Edu)
```{r}
UNdataExercise[UNdataExercise$Edu == max(UNdataExercise$Edu),'HDI']
```

Exercise 5: Use a single command to list the continents (each one appearing once)
```{r}
unique(UNdataExercise$continent)
```

Exercise 6: Use a single command to return the average population of the country belonging to the continent of Asia
```{r}
mean(UNdataExercise[UNdataExercise$continent == 'Asia','Population'])
```
