---
title: "ggplot advanced"
output: html_document
---
## ggplot from the ggplot2 package
### Advanced ggplot
In this rmd file we explore advanced ggplot functionality, including:
* themes
* annotations
* dates
* maps
* predictions
* interactivity (plot.ly)

We will use a dataset of college metrics.
```{r load packages and data}
library(ggplot2)
library(ggthemes)
library(scales)
library(plotly)

college2.df <- read.csv('college2.csv')
head(college2.df) 
str(college2.df) 
dim(college2.df) 
```

In addition to the data selection, aesthetic mapping, geom selection, and selection of other parameters, the creator of a ggplot has a wide range of options for formatiing. While this formatting can be done piecemeal, that is, a single selection at a time, the more common appraoch is the creation and use of a theme. A theme consists of the naming of a series of generic formatting options that are then applied holistically to a plot.
```{r themes}
my_theme <- theme(
  plot.title=element_text(size=20, color = 'pink', margin = margin(10, 10, 10, 10), face="bold"),
  axis.text.x=element_text(size=15, color = 'coral1'),
  axis.text.y=element_text(size=15, color = 'cyan1'),
  axis.title.x=element_text(size=15, color = 'coral1'),
  axis.title.y=element_text(size=15, color = 'cyan1'),
  panel.background = element_rect(fill = 'grey15'),
  legend.title = element_text(color='green', size=10, face='bold')
  ) 
  
# The following two commands create a ggplot object and render the plot
college.p <- ggplot(data = college2.df, aes(x = application.count, y = acceptance.count/application.count, color = private))
college.p + geom_point()

# We first leverage a built in theme function: theme_bw() 
college.p + geom_point() + theme_bw() # built in theme

# We now use a custom theme; not this theme is a variable of type list rather than a theme function. The two additions, ggtitle and the labs values, use the theme formatting settings
college.p + geom_point() + my_theme + 
  ggtitle('College Prestige') + 
  labs(x = 'Applications', y = 'Acceptance Rate') # Using labels parameter

# The 'last' theme's formatting takes precedent in case of conflict as shown below. Alternatively, narrowly defined themes (e.g., one for axis, one for legends, etc.) can be used in combination.
college.p + geom_point() + my_theme + theme_bw() + 
  ggtitle('College Prestige') + 
  labs(x = 'Applications', y = 'Acceptance Rate') # Using labels parameter
```

```{r other}
college.p <- ggplot(data = college2.df, aes(x = application.count, y = acceptance.count/application.count, color = private))

# Using scale_x_continuous instead of labs parameter appears to be the more commonly used approach for creating axis names 
college.p + geom_point() + my_theme + ggtitle('College Prestige') +
  scale_x_continuous(name = "Applications") +
  scale_y_continuous(name = "Acceptance Rate")

# The reason for this is that these layers accept several parameters in addition the the axis name such as axis ranges, tick ranges, and formatting
college.p + geom_point() + my_theme + ggtitle('College Prestige') +
  scale_x_continuous(limits=c(0, 30000), breaks=c(0, 15000, 30000), name = "Applications") +
  scale_y_continuous(name = "Acceptance Rate", labels = percent)

# Rather than set a range (via the limits parameter), the coor_cartesian 'zooms' into the referenced part of the plot. This may be important as the former drops data points while the latter does not, potentially affecting some stat calculations 
college.p + geom_point() + my_theme + ggtitle('College Prestige') +
  scale_x_continuous(breaks=c(0, 15000, 30000), name = "Applications") +
  scale_y_continuous(name = "Acceptance Rate", labels = percent) +
  coord_cartesian(xlim=c(0,20000)) # zooms instead of truncate values

# We next rename the legend
college.p + geom_point() + my_theme + ggtitle('College Prestige') + 
  scale_x_continuous(name = "Applications") +
  scale_y_continuous(name = "Acceptance Rate", labels = percent) +
  scale_color_discrete(name="Private School") # rename legend label

# If the legend was built using a size aesthetic instead of color, use scale_size("Name") to rename legend or scale_size(range = c(10, 100)) to adjust glyph sizes

# Others can provide formatting options through their creation of themes, such as through the ggthemes package (in this case, using theme_economist)
college.p + geom_point() + 
  theme_economist() + 
  ggtitle('College Prestige') + 
  scale_y_continuous(name = "Acceptance Rate", labels = percent) +
  scale_color_discrete(name="Private School") 

# We can override formatting by declaring a single formatting option that overrides the theme as well (in this case, color)
college.p + geom_point() + 
  theme_economist() + 
  ggtitle('College Prestige') + 
  scale_y_continuous(name = "Acceptance Rate", labels = percent) +
  scale_color_discrete(name="Private School") +
  scale_color_manual(values=c("darkgreen", "goldenrod1")) # override color aesthetics
```

```{r custom color palettes}
college.p <- ggplot(data = college2.df, aes(y =  phd.granted.count, x = s.f.ratio.factor, fill = s.f.ratio.factor)) + geom_boxplot()
college.p

college.p + 
    scale_fill_hue(c=45, l=80)

college.p + 
    scale_fill_brewer(palette="Spectral")

college.p + 
    scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99", "#66CC99", "#66CC99"))

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
college.p + 
 scale_fill_manual(values=cbPalette)

# To use custom colors for line and point colors, add scale_colour_manual(values=cbPalette)
  
```

ggplot allows intricate annotation options beyond the simple placement of a text field. In the example below, we highlight and label the 'sweet spot' of the plot, that is, the subjectively selected desirable x and y coordinates that represent 'success' on the part of the college.
```{r annotations}
college.p <- ggplot(data = college2.df, aes(x = application.count, y = acceptance.count/application.count, color = private))

college.p + geom_point() + 
  annotate("text", label = "Target grid", x = 17000, y = .25, size = 3, fontface = "italic") +
  annotate("rect", xmin = 10000, xmax = 15000, ymin = .1, ymax = .25, alpha = .2, color = 'red', fill = 'coral1') + 
  annotate("segment", x = 14000, xend = 17000, y = .18, yend = .25, color = "red")
  
```

Date manipulation and calculations in R often require their own set of functions. However, the plotting of dates in ggplot is straightforward. 
```{r dates}
sept <- read.csv('temps.csv')
head(sept) # date in mm/dd/yy format
sept$date2 <- strptime(sept$date, "%m/%d/%y")
head(sept) 
str(sept)
sept.p <- ggplot(data = sept, aes(x = date2))
sept.p  + 
  geom_smooth(aes(y = high), se = FALSE, color = 'coral1') + 
  geom_smooth(aes(y = low), se = FALSE, color = 'cyan1') +
  geom_smooth(aes(y = ((high + low)/2)), se = FALSE, color = 'goldenrod3', lwd = .5, linetype = 'dotdash') +
  geom_hline(aes(yintercept = mean(high)), size = 1.5, color = 'coral3') +
  geom_hline(aes(yintercept = mean(low)), size = 1.5, color = 'cyan3') +  scale_y_continuous(name = 'Temperatures') +
  scale_x_datetime(name = 'Dates') # Note the reference to the x_datetime axis instead of x_continuous

```


```{r plotly}
college.p <- ggplot(data = college2.df, aes(x = application.count, y = acceptance.count/application.count, color = private))

college.p + geom_point()
ggplotly() # turns plot into JSON; can filter via legend, hoover over point for pop-up, zoom in via click and drag on grid

```

### Note 1
For 3d graphs, consider the lattice package. For network graphs, consider the igraph package.
